Research Report: Longitudinal models of human mobility

### 2017-11-10 thesisReport.Rmd
added graph plus methods and abstract bits

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggthemes)
library(ggplot2)
```

@@

Longitudinal data on human temporospatial behaviour has become available through smartphones and other sensors. This data has been used to investigate topics such as the effects of the food environment on eating patterns *Zenk*, the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017]. While there have been advances in processing this data [@zheng_mining_2009] extracting meaningful information remains a difficult task. Two important challenges are dealing with measurement noise and missing data.

With regards to measurement accuracy, studies of professional grade GPS trackers suggest that less than 80% of measurements fall within 10 meters of the true location. Moreover, GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this happens to be where most people in the developed world tend to spend most of their time. In addition, location data collected by more ubiquitous (but less specialised) smartphones are an amalgamation of sensor data. For instance, Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006]. This adds another layer of complexity as each of these measures has its own characteristics in terms of measurement accuracy and bias.

@@

There is currently no golden standard in how to deal with missing temporospatial data [@barnett_inferring_2016]. @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it. Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017] can increase the severity of the missing data problem. On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for.

In this paper we will compare methods used to deal with measurement error and missing data in location. Specifically, we are interested in location tracks as measured by commerically available smartphones.

## State of the art

Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and often serves different purposes. The following section briefly illustrates the different methods used to solve the measurement inaccuracy and missing data problem.

@@

There is a vast literature of using state space models (SSMs) to improve measurements accuracy and deal with missing data. Behavioural ecologists for instance, have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex, for example *Priesler* uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. *Gurram* *Kalman* *Chen* *Brown*. The extended Kalman filter is the de facto standard for GPS measurements.

@@

For highly correlated in time.

@@

![Raw GPS measurements of a journey from de Uithof to Tuinwijk on February 17th 2017. The measurements are in red, the filtered path is in blue. The circles denote 67% confidence intervals of the given GPS measurement. Measurements and fitted points which follow each other in time are connected by lines. The inacurate measurements lead to estimates of irregular movements. The filtered movement estimate seems more accurate in terms of the path, but lags behind the unfiltered movement.](//img//Map1.png)

@@

ps<- readRDS("..//data//missingDataAnalysisBoaz.rds")

pal<- c("#4228A8", "#0D0101","#E04AC2")

ggplot(ps, aes(x = Day, y=count, fill = type, colour= type))+geom_point()+
  theme_tufte()+
  xlab("")+
  ylab("Measurements")+
  ggtitle("Missing data over time")+
  scale_colour_manual(values=pal)+
  theme(legend.position = c(0.8, 0.8))+
  theme(legend.title=element_blank())

rm(ps)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
```

# Abstract
*There is a great interest in using objective location data to quantify the mobility characteristics of individuals. However, *

@@

Objective data on human temporospatial behaviour has become available throgh the Global Positioning System (GPS) which uses the distance between a device and a number of satellites to determine location. Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality and academic performance [@harari_using_2016; @wang_smartgpa:_2015] and detecting bipolar disorder [@palmius_detecting_2017]. In most of these studies participants are given a specialised GPS devices or a smartphone with a custom made app. However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants, moreover may be biased because of the introduction of a new device to the participant's life and usually span a short amount of time. A solution would be to take advantage of the location logs that smartphones already collect, such as Google Location History which stores Android phone users' locations [@location_history_timeline_nodate]. Yet, because GPS sensors consume a significant amounts of battery, these logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data.

Studies of professional grade GPS trackers suggest that less than 80% of measurements fall within 10 meters of the true location. Moreover, GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for social scientists , this happens to be where most people in the developed world tend to spend most of their time. In addition, location data collected by more ubiquitous (but less specialised) smartphones are an amalgamation of sensor data. For instance, Android phones collect location information from WiFi access points, cellphone triangluation, and GPS measurements due to computational and battery constraints [@lamarca_place_2005;@chen_practical_2006]. This adds another layer of complexity as each of these measures has its own characteristics in terms of measurement accuracy and bias.

@@

There is currently no golden standard in how to deal with missing temporospatial track data [@barnett_inferring_2016]. @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with it. Methods frequently used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017] can increase the severity of the missing data problem. On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for.

In this paper we will compare methods used to deal with measurement error and missing data in location information. Specifically, we are interested in establishing accurate mobility patterns from smartphone GPS logs.

# Problem description and literature review

## The structure of smartphone location logs

Android logged data

```{r, measurementsPerDay,echo = FALSE,fig.cap="Missing data over time for the author. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a day. The fill of the points shows the amount of hours of captured data each day.",out.width = '100%'}
include_graphics("img/missingdayPeter.png")# from the playingaroundwiththicken.r file
```

Linear Interpolation vs actual movement vs kalman filtered.

## State of the art in temporospatial models

Research with respect to the analysis of GPS data is wideranging, highly interdisciplinary and often serves different purposes. Given that @barnett_inferring_2016 is to our knowledge the only paper explicitly focusing on dealing with missingness for inferring missingness using GPS data the following section briefly illustrates the some methods used to solve measurement inaccuracy and missing data problems in the field, as well as their applicability to our question.

Given fixed measurement stations there are sevaral imputation methods for spatiotemporal measurements. For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.

While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostentibly by dividing the sample space into rasters which would be measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible to do. To our knowledge such models have not been implemented.

@@

There is a vast literature of using state space models (SSMs) to improve measurements accuracy and deal with missing data. Behavioural ecologists for instance, have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. The extended Kalman filter is the de facto standard for GPS measurements [@chen_state_2013].

@@

@@

![Raw GPS measurements of a journey from de Uithof to Tuinwijk on February 17th 2017. The measurements are in red, the filtered path is in blue. The circles denote 67% confidence intervals of the given GPS measurement. Measurements and fitted points which follow each other in time are connected by lines. The inacurate measurements lead to estimates of irregular movements. The filtered movement estimate seems more accurate in terms of the path, but lags behind the unfiltered movement.](img/Map1.png)

@@
